---
title: Analysis - Covid with other factors influencing unemployment
  rate
author: "Jaya Johnson"
date: "4/19/2021"
output: pdf_document
---
**Summary of Proposed Analysis**. 
In this section we are analyzing the impact of various factors that may or may not contribute to the overall unemployment rate. Some of the contributing factors are: 

1) Race (we are taking into consideration the unemployment rate of African Americans, Hispanic Americans, White Americans, Asian Americans).   

2) Gender (unemployment rate for males and females).  

3) Overall monthly job loss percent.   

4) Monthly Covid Numbers.  

As a part of this analysis we would like to understand the relationship between the different contributing factors and employment rate.  

We would take the most influential parameters, come up with different hypothesis and come up with a multiple regression analysis to predict the unemployment rate.

**Data Analysis**


```{r}
library(readr)
Covid_and_Unemployment_US <- read_csv("Covid and Unemployment US - Sheet1.csv")
```

```{r}
unemployment_df <- data.frame(Covid_and_Unemployment_US)
unemployment_df[is.na(unemployment_df)] = 0
unemployment_df$Date <- as.Date(unemployment_df$Date , format = "%d/%m/%y")
head(unemployment_df)

```
```{r}
library(funModeling) 
library(tidyverse) 
library(Hmisc)
```
```{r}
print("Raw Data Metrics")
status(unemployment_df)
```
```{r}
plot_num(unemployment_df)
```


```{r}
library(ggplot2)
library(dplyr)
```

```{r}
library(cowplot)
gp1<-ggplot(data = unemployment_df, aes(x = Date, y = Unemployment.Rate)) + 
  geom_line(color = "#fc03f0", size = 1) +
  xlab("") +
  ylab("Unemployment Rate")+
  theme(
  axis.title.y = element_text(color="blue", size=6, face="bold"),
)
gp2<-ggplot(data = unemployment_df, aes(x = Date, y = Unemployment.Rate.AfricanAmerican)) +
  geom_line(color = "#FC4E07", size = 1) +
  xlab("") +
  ylab("UnemploymentRate - African Americans")+
  theme(
  axis.title.y = element_text(color="blue", size=6, face="bold"),
)
gp3<-ggplot(data = unemployment_df, aes(x = Date, y = Unemployment.Rate.White)) + 
  geom_line(color = "#8403fc", size = 1) +
  xlab("") +
  ylab("UnemploymentRate - White")+
  theme(
  axis.title.y = element_text(color="blue", size=6, face="bold"),
)
gp4<-ggplot(data = unemployment_df, aes(x = Date, y = Unemployment.Rate.Asian)) + 
  geom_line(color = "#036ffc", size = 1) +
  xlab("") +
  ylab("UnemploymentRate - Asian")+
  theme(
  axis.title.y = element_text(color="blue", size=6, face="bold"),
)
gp5<-ggplot(data = unemployment_df, aes(x = Date, y = Unemployment.Rate.Hispanic)) + 
  geom_line(color = "#03e7fc", size = 1) +
  xlab("") +
  ylab("UnemploymentRate - Hispanic")+
  theme(
  axis.title.y = element_text(color="blue", size=6, face="bold"),
)
gp6<-ggplot(data = unemployment_df, aes(x = Date, y = Unemployment.Rate.Men)) + 
  geom_line(color = "#03fcd3", size = 1) +
  xlab("") +
  ylab("UnemploymentRate - Male")+
  theme(
  axis.title.y = element_text(color="blue", size=6, face="bold"),
)
gp7<-ggplot(data = unemployment_df, aes(x = Date, y = Unemployment.Rate.Women)) + 
  geom_line(color = "#98fc03", size = 1) +
  xlab("") +
  ylab("UnemploymentRate - Female")+
  theme(
  axis.title.y = element_text(color="blue", size=6, face="bold"),
)
gp8<-ggplot(data = unemployment_df, aes(x = Date, y = 	Percent.JobLosers)) + 
  geom_line(color = "#fce303", size = 1) +
  xlab("") +
  ylab("Percent of Job Losers")+
  theme(
  axis.title.y = element_text(color="blue", size=6, face="bold"),
)
gp9<-ggplot(data = unemployment_df, aes(x = Date, y = 	CovidCases)) + 
  geom_line(color = "#fc6f03", size = 1) +
  xlab("") +
  ylab("Covid Cases")+
  theme(
  axis.title.y = element_text(color="blue", size=6, face="bold"),
)
plot_grid(gp1, gp2, gp3, gp4, gp5, gp6, gp7, gp8, gp9, labels = c('','','','','','','','',''), label_size = 4)
```

```{r}
library(viridis)
gp1<-ggplot(data = unemployment_df, aes(x=Unemployment.Rate, y=Unemployment.Rate.AfricanAmerican)) + 
geom_point(aes(colour=Unemployment.Rate.AfricanAmerican))+
scale_color_viridis(option = "D")

gp1

gp2<-ggplot(data = unemployment_df, aes(x=Unemployment.Rate, y=Unemployment.Rate.Asian)) + 
geom_point(aes(colour=Unemployment.Rate.Asian))+
scale_color_viridis(option = "A")

gp2

gp3<-ggplot(data = unemployment_df, aes(x=Unemployment.Rate, y=Unemployment.Rate.White)) + 
geom_point(aes(colour=Unemployment.Rate.White))+
scale_color_viridis(option = "B")

gp3

gp4<-ggplot(data = unemployment_df, aes(x=Unemployment.Rate, y=Unemployment.Rate.Hispanic)) + 
geom_point(aes(colour=Unemployment.Rate.Hispanic))+
scale_color_viridis(option = "C")

gp4

gp5<-ggplot(data = unemployment_df, aes(x=Unemployment.Rate, 
y=Unemployment.Rate.Men)) +
geom_point(aes(colour=Unemployment.Rate.Men))+
scale_color_viridis(option = "D")

gp5

gp6<-ggplot(data = unemployment_df, aes(x=Unemployment.Rate, y=Unemployment.Rate.Women))+
geom_point(aes(colour=Unemployment.Rate.Women))+
scale_color_viridis(option = "E")

gp6

gp7<-ggplot(data = unemployment_df, aes(x=Unemployment.Rate, 
y=Percent.JobLosers))+
geom_point(aes(colour=Percent.JobLosers))+
scale_color_viridis(option = "F")

gp7

gp8<-ggplot(data = unemployment_df, aes(x=Unemployment.Rate, y=CovidCases))+
geom_point(aes(colour=CovidCases))+
scale_color_viridis(option = "G")

gp8

```

